Composable Concurrency Models

### 2016-11-23 paper/paper.tex
Cleaned up presentation intro/background and typeset diagrams
Made some changes for the final draft

\documentclass{sig-alternate}
\usepackage[USenglish]{babel}
\usepackage[backend=bibtex,style=numeric-comp]{biblatex}
\usepackage{graphicx}

\graphicspath{{../figures/}}

@@

The need to manage concurrent operations in applications has led to the development of a variety of concurrency models. Modern programming languages generally provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem. We discuss challenges surrounding concurrency control and examine situations in which conflicts between models can occur. Additionally, we describe attempts to implement concurrency models on top of lower-level concurrency abstractions.

@@

\keywords{concurrency, parallelism, concurrency abstraction}

@@

Most interactive computer programs depend on concurrency, the ability to perform different tasks at the same time. A web browser, for instance, might at any point be rendering documents in multiple tabs, transferring files, and handling user interaction. On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events~\cite{Swalens2014}. If every long-running process blocked other processes from proceeding, the system would be effectively unusable.

Concurrency models enable programmers to reason about and describe interactions between concurrent processes. The web browser is a good example of several different models: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations to achieve acceptable performance~\cite{Marr2012}.

Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could be arbitrarily combined. From an engineering standpoint, composable models are also desir\-able---a virtual machine for one or more high-level languages should be able to support a variety of models without resorting to rough low-level approximations. However, different concurrency models are not necessarily composable and unexpected issues may arise when they interact. Recent work has attempted to identify unifying concurrency abstractions that would guarantee composability of different models and also provide underlying virtual machine support~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.

@@

Modern operating systems are expected to run many processes concurrently, and processes themselves are often composed of multiple concurrent threads of execution. A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}. Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.

Concurrency models can abstract over implementation details, allowing programmers to reason in terms of asynchron\-ous tasks and independent components instead of low-level thread management. In addition to making a program more understandable, higher-level models afford a degree of flexibility: Regardless of whether a set of concurrent operations is executed in sequence on one thread, in multiple threads on the same processor, or in parallel on multiple processors, the result will be the same.

\subsection{Concurrency}

For operations to be executed concurrently, they must be able to be executed out of order or in partial order. Lamport, in his foundational work on distributed systems~\cite{Lamport1978}, formalized this by defining a ``happens before'' relation (denoted by ``$\rightarrow$'' and its negation by ``$\nrightarrow$'') on a set of operations: If $A$ and $B$ are operations in the same process and $A$ occurs before $B$, or if $A$ is the sending of a message by one process and $B$ is the receipt of the same message by another process, then $A \rightarrow B$. Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.

While the ``happens before'' relation can be used to determine whether operations can be executed concurrently, it does not guarantee the correctness of the results. In a non-concurrent (entirely sequential) program, it would be sufficient to show that the program is correct after the execution of each operation and therefore correct upon termination. In other words, the program could be proved to be correct by proving that its history (that is, the sequence in which its operations are performed) yields a correct result. Consistency models guarantee the correctness of a program by defining a set of all allowed histories, or histories that lead to a correct result~\cite{Ziv2015}. If an execution of a program follows an allowed history, the execution is consistent. If not, the execution is inconsistent. If every possible execution of the program follows an allowed history, the program conforms to the consistency model.

\begin{figure}[h]
  \centering
  \includegraphics{single-thread-read-write}
  \caption{A single thread writes to and reads from a variable.}
\label{figure:single}
\end{figure}

Figure~\ref{figure:single} illustrates a simple program in which a single thread reads and writes numbers from a variable $x$. The value of $x$ over time is represented by a segmented bar. Write operations are denoted by an arrow pointing toward the bar and read operations by an arrow pointing away. The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value is equal to the value of the most recent write.

\begin{figure}[h]
  \centering
  \includegraphics{multiple-threads-read-write}
  \caption{Two threads concurrently write to and read from a variable.}
\label{figure:multiple}
\end{figure}

In a concurrent program, the history of operations may not be the same for every execution. As a result, assumptions about correctness that are true for a sequential program may be violated in a concurrent program. This is illustrated in Figure~\ref{figure:multiple}, in which two threads concurrently write from and read to a variable $x$ concurrently. Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write. The value may even be inconsistent between consecutive reads. If a single thread were to incorrectly assume that it had exclusive control of $x$, this behavior would appear incorrect and could lead to race conditions, errors that arise due to unintentional dependency on execution order.

\subsection{Linearizability}

@@

\begin{figure}[h]
  \centering
  \includegraphics{noninstantaneity}
  \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
\label{figure:time}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics{linearizability}
  \caption{A single thread performs an atomic write operation on a variable.}
\label{figure:linearizability}
\end{figure}

@@

\subsection{Serializability}

@@

\begin{figure}[h]
  \centering
  \includegraphics{serializability}
  \caption{Two threads perform a set of serializable read and write operations on a variable.}
\label{figure:serializability}
\end{figure}

Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations. Even though the exeuctions of each operation occur out of order and even intersect, the events themselves can be arranged serially: $x \leftarrow 1$, $x = 1$, $x \leftarrow 2$, $x = 2$.

Serializability alone is a fairly weak consistency model. Because it does not place bounds on time or order, events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee. Serializability is useful, though, as a guarantee of isolation~\cite{Haerder1983}: While a serializable set of operations is being executed, it appears to be the only set of operations being executed. For histories to be serializable, then, they must not overlap and cannot interfere with each other.

Together, linearizability and serializability guarantee strict serializability, in which a history is equivalent to a serial execution order and that serial order corresponds to the execution order in real time. Given that, linearizability can actually be described as strict serializability restricted to single operations on single objects~\cite{Herlihy1990}.

@@

Modern programming languages generally provide several different concurrency models, each with their own terminology and specific implementation details. It would be impractical to discuss them all, but the most commonly used models can be adequately described by generalization~\cite{Swalens2014}.

@@

Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible. \code{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.

@@

Many higher-level languages (such as Clojure~\cite{Swalens2014}) provide atomic references that can be read by an atomic dereferencing operation and written by an atomic swap operation. Such operations may provide convenience features such as automatic retries.

@@

Software transactional memory (STM) presents an alternative to semaphores and locks by allowing multiple concurrent operations to write to a shared location in memory. STM is an example of optimistic concurrency control: Each thread executes a series of read and write operations (called a transaction) on the shared memory without considering the activity of other threads. On its face, this approach seems disaster-prone---multiple threads writing to the same memory location could easily lead to corrupted data. STM solves this problem by recording all operations in a log. After the entire transaction is completed, the transaction manager verifies that other threads have not also made changes to the shared memory. If there are conflicting changes, the transaction is aborted and retried until it eventually succeeds~\cite{Shavit1995}.

Conceptually, STM simplifies concurrency because it allows a transaction to be thought of as a single strictly serializable operation. A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed. Only when a transaction completes successfully will changes will become visible to other threads. Practically, STM is useful in any situation in which a shared object needs to be accessed and modified by multiple threads~\cite{Swalens2014}.

STM also offers significant advantages over lock-based programming in terms of composability. Given that executing two operations within a transaction yields a larger atomic operation, Harris et al.\ describe a STM that allows any combination of atomic operations to be composed into a larger atomic operation~\cite{Harris2005}.

However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations. Specifically, a transaction cannot execute any operation that cannot be undone or retried, such as writing to disk or performing a network request. STM can also result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.

@@

Other concurrently models avoid the use of shared memory (and the problems that arise from doing so) by restricting threads to private memory. Threads communicate strictly by message passing, which avoids race conditions. A variety of different communicating thread models exist and have long been a solution to describing concurrent operations. Practical use cases for communicating threads include communication with external systems or running an event loop~\cite{Swalens2014}.

Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels~\cite{Hoare1978}. CSP is notable in that messages are passed synchronously; that is, a sending process will block until a complementary operation is executed on the receiving process~\cite{Swalens2014}.

Other models, most notably the actor model, rely on asynchronous message passing to specific entities~\cite{Agha1986}. Actors can only communicate via asynchronous message passing. Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.

@@

Futures and promises are two of the most common proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual''). Generally, futures are resolved to the result of the completed operation. The result is then accessed implicitly; any use of the future will return its value. Promises, on the other hand, are created as independent objects and require the result to be accessed explicitly.

In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations. Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks~\cite{Kambona2013}.

\section{Proposed abstractions}

\subsection{Asynchronous events}

Ziarek et al.~\cite{Ziarek2011} introduce asynchronous events as a possible abstraction. Their examples are heavily tied to Concurrent ML and might be difficult to integrate.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. To identify concepts relevant for a multi-language virtual machine, Marr and D'Hondt selected four questions:

\begin{enumerate}
  \item Can the concept be reasonably implemented as a library? (A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant virtual machine inclusion.)

  \item Does the concept require runtime support to guarantee its semantics? (Semantic guarantees may be enforced by a compiler but not the virtual machine. For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.)

  \item Would runtime support benefit performance? (Deeper knowledge of the concept may enable the virtual machine to better optimize. Some concepts, though, may require knowledge of the underlying hardware to realize any performance improvements; such concepts would not necessarily benefit from runtime support.)

@@

\begin{itemize}
  \item Managed mutation: Concurrency models impose restrictions on how state can be modified; therefore, mutation must be able to be flexibly managed.

@@

  \item Ownership: Some owning entity usually regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

  \item Leveled reflection: Reflection allows a program to examine its own structure and behavior at runtime. To guarantee safety, there needs to be a distinction between language-level reflection (restricted reflection operations supported by the high-level language) and meta-level reflection (unrestricted reflection operations on the low-level abstraction).

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=224pt]{obmop}
  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

@@

In more words: There's a lot of work on the correctness of concurrent operations. With some care, concurrency models can be composed. There's some work on deeper abstractions, but it's definitely ongoing and more is likely to follow soon.

@@

Thanks to Elena Machkasova and K.~K. Lamberty for their patience and feedback.

>>>>

\documentclass{sig-alternate}
\usepackage[USenglish]{babel}
\usepackage[backend=bibtex,style=numeric-comp]{biblatex}
\usepackage{tikz}
\usetikzlibrary{positioning}

@@

The need to manage concurrent operations in applications has led to the development of a variety of concurrency models. Modern programming languages generally provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem. We discuss challenges surrounding concurrent programming and examine situations in which conflicts between models can occur. Additionally, we describe attempts to identify features of common concurrency models and develop lower-level abstractions capable of supporting a variety of models.

@@

@@

Most interactive computer programs depend on concurrency, the ability to perform different tasks at the same time. A web browser, for instance, might at any point be rendering documents in multiple tabs, transferring files, and handling user interaction. On a lower level, the operating system might be running several other applications, juggling background processes, and responding to events. If every long-running process blocked other processes from proceeding, the system would be effectively unusable.

Processes themselves are often composed of multiple concurrent threads of execution. A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}. Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.

Concurrency models enable programmers to reason about concurrent tasks instead of low-level thread management. The web browser is a good example of how different models might be chosen to represent specific types of tasks: The user interface layer might rely on an event loop, the rendering process might operate in shared memory, and suggestions from browsing history or a search engine might require parallel collection operations to achieve acceptable performance~\cite{Marr2012}.

Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could safely interact. However, different models do not necessarily work well together, nor are they designed to. Recent work has attempted to identify common ``building blocks'' that could be used to compose a variety of models, possibly eliminating subtle problems when different models interact and allowing models to be represented at lower levels without resorting to rough approximations~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.

@@

In a concurrent program, the history of operations may not be the same for every execution. An entirely sequential program could be proved to be correct by showing that its history (that is, the sequence in which its operations are performed) always yields a correct result. For a concurrent program to be proved correct, though, it must be clear that all possible histories always yield a correct result. Furthermore, concurrent programming frequently involves the use of shared resources that require coordinated access and manipulation.

Figure~\ref{figure:single} illustrates a simple program in which a single thread reads ($=$) and writes ($\leftarrow$) values on a variable $x$, denoted by a segmented bar. The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value returned is equal to the value most recently written. Because the thread has exclusive control of $x$, that assumption is valid.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
  \caption{A single thread reads from and writes to a variable.}
\label{figure:single}
\end{figure}

In Figure~\ref{figure:multiple}, on the other hand, two threads concurrently read and write on $x$. Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write. The value may even be inconsistent between consecutive reads. If a single thread incorrectly assumed that it had exclusive control of $x$, this behavior could lead to inconsistent or incorrect results. Similar errors that arise due to an unintentional dependency on execution order are commonly referred to as race conditions.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
  \caption{Two threads concurrently read from and write to a variable.}
\label{figure:multiple}
\end{figure}

Consistency models guarantee the correctness of a program by defining a set of all allowed histories, or histories that lead to a correct result~\cite{Ziv2015}. If an execution of a program follows an allowed history, the execution is consistent. If not, the execution is inconsistent. If every possible execution of the program follows an allowed history, the program conforms to the consistency model.

\subsection{Linearizable consistency}

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/noninstantaneity.tikz}}
  \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
\label{figure:time}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/linearizability.tikz}}
  \caption{A single thread performs an atomic write operation on a variable.}
\label{figure:linearizability}
\end{figure}

@@

\subsection{Sequential and causal consistency}

For operations to be executed concurrently, they must be able to be executed out of order or in partial order. Lamport, in his foundational work on distributed systems~\cite{Lamport1978}, formalized this by defining a ``happens before'' relation ($\rightarrow$) and its negation ``does not happen before'' ($\nrightarrow$) on a set of operations: If $A$ and $B$ are operations in the same thread and $A$ occurs before $B$, or if $A$ is the sending of a message by one thread and $B$ is the receipt of the same message by another thread, then $A \rightarrow B$. If $A \rightarrow B$ and $B \rightarrow C$, then $A \rightarrow C$. Two operations $A$ and $B$ are said to be concurrent if $A \nrightarrow B$ and $B \nrightarrow A$.

The first condition of the ``happens before'' relation, that $A$ and $B$ are ordered within the same thread, describes sequential consistency. Sequential consistency simply requires that operations in a thread take place in that thread's order. Causal consistency, described by the second condition of the ``happens before'' relation, enforces the intuitive requirement that an effect cannot take place before its cause.

\subsection{Serializable consistency}

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/serializability.tikz}}
  \caption{Two threads perform a set of serializable read and write operations on a variable.}
\label{figure:serializability}
\end{figure}

Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations. Even though the executions of each operation occur out of order and even intersect, the events themselves can be arranged serially.

Serializability alone is a fairly weak consistency model. Because it does not place bounds on time or order, events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee. Serializability is useful, though, as a guarantee of isolation: While a serializable set of operations is being executed, it appears to be the only set of operations being executed. For histories to be serializable, then, they must not overlap and cannot interfere with each other.

Linearizability and serializability together guarantee strict serializability, in which a history is equivalent to a serial execution order and that serial order corresponds to the execution order in real time. Given that, linearizability can actually be described as strict serializability restricted to single operations on single objects~\cite{Herlihy1990}.

@@

Modern programming languages generally provide several different concurrency models as tools to reason about concurrent tasks. Swalens et.\ al, in a survey of concurrency models in the Clojure programming language, group commonly used concurrency models into four broad categories: atomic variables, software transactional memory, communicating threads, and proxies~\cite{Swalens2014}.

@@

Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible. \code{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.

@@

While lock-based programming is effective, dealing with multiple locks and complex interactions can be cumbersome. One common problem in lock-based programming is resource contention, in which multiple threads vie for control of multiple resources. Suppose, for example, that two threads $A$ and $B$ both require resources $x$ and $y$. $A$ might acquire $x$ and $B$ might acquire $y$. Both would then halt, waiting to acquire a lock on the second needed resource. This type of scenario is commonly referred to as deadlock.

@@

Software transactional memory (STM) presents an alternative to lock-based programming by allowing multiple concurrent operations to write to a shared location in memory without securing any kind of lock. STM is an example of optimistic concurrency control: Each thread executes a series of read and write operations (called a transaction) on the shared memory without considering the activity of other threads. On its face, this approach seems disaster-prone---multiple threads writing to the same memory location could easily lead to corrupted data. STM solves this problem by recording all operations in a log. After the entire transaction is completed, the transaction manager verifies that other threads have not also made changes to the shared memory. If there are conflicting changes, the transaction is aborted and retried until it eventually succeeds~\cite{Shavit1995}.

Conceptually, STM simplifies concurrency because it allows a transaction to be thought of as a single strictly serializable operation. A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed. Only when a transaction completes successfully will changes will become visible to other threads. Wrapping a set of operations within a transaction eliminates the need to secure locks, leading to more simple programs. Practically, STM is useful in any situation in which a shared object needs to be accessed and modified by multiple threads~\cite{Swalens2014}.

STM also offers significant advantages over lock-based programming in terms of composability. Because a transaction can be viewed as a single linearizable operation, STM can allow any combination of linearizable operations to be composed into a larger linearizable operation. Transactions also enforce modularity by hiding implementation details of the operations within~\cite{Harris2005}.

However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations. Specifically, a transaction cannot execute any operation that cannot be undone or retried, such as writing to disk or performing a network request. Operations that block indefinitely may also cause a transaction to fail continuously. STM can result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.

@@

Other concurrently models avoid the use of shared memory by restricting threads to private memory. Threads communicate strictly by message passing, which avoids issues such as race conditions. Sending messages and reacting to received messages are inherently asynchronous processes, so communicating threads have long been a solution to describing concurrent operations. Limiting thread communication to message passing also enforces modularity and limits possible negative side effects. A wide variety of communicating thread models exist and often run event loops or handle communication with external systems~\cite{Swalens2014}.

Communicating sequential processes (CSP), one of the oldest communicating thread models, began as a method of formally describing actions in concurrent systems. CSP describes systems in terms of independent processes that communicate through predefined channels~\cite{Hoare1978}. CSP is notable in that messages are passed synchronously; that is, a sending process will block until a complementary operation is executed on the receiving process~\cite{Swalens2014}. CSP is still widely used in the specification, implementation, and testing of safety-critical systems.

Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}. Actors can only communicate via asynchronous message passing. Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.

@@

Futures and promises are two of the most common proxy models. The terms are frequently used interchangeably, along with ``delayed,'' ``deferred,'' and ``eventual.'' Generally, futures are resolved to the result of the completed operation. The result is then accessed implicitly; any use of the future will return its value. Promises differ in that they are created as independent objects and require the result to be accessed explicitly.

In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations. Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks.

\section{Composability challenges}

While concurrency models guarantee their own correctness, issues may arise when different models are used together. For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}. If such a resource is modified outside of the STM, an unexpected interleaving of operations could result.

Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models. The study surveyed all of the concurrency models available in the Clojure programming language and used them in tandem with one another to see what types of correctness issues were encountered.

\subsection{Correctness criteria}

In the study, safety and liveness were used as the two criteria for evaluating the correctness of a combination of models. Together, the two have historically been used to prove the correctness of distributed systems~\cite{Lamport1977}.

Informally, safety guarantees that ``nothing bad will happen.'' In other words, given a correct input, a program will not produce an incorrect result~\cite{Swalens2014}. This property is also referred to as partial correctness. Safety is generally achieved in the management of shared resources. STM, for instance, only allows shared memory to be accessed through transactions; communicating threads only allow data to be shared through message passing.

Liveness guarantees that ``something good will eventaully happen,'' or that a program will eventually terminate if its input is correct. Taken together, safety and liveness are referred to as total correctness---given a correct input, a program will terminate with the correct output~\cite{Swalens2014}. Deadlocks and livelocks are the primary obstacle to liveness and can arise when models are combined.

\subsection{Possible conflicts}

To discover the types of conflicts that might arise when combining different models, the study examined all pairwise combinations of Clojure's models. Using every model within every other model uncovered the following types of conflicts:

\begin{itemize}
  \item A model might reexecute code containing another concurrency model that performs an irrevocable action. For instance, if a transaction were to contain an operation that sent a message on a channel, reexecutions of the transaction would cause the message to be sent repeatedly.

  \item A model might reexecute code that causes the reexecution to continually happen. The authors of the study observed that multiple STM transactions might consistently conflict with another and never succeed.

  \item A model that supports blocking operations might be used within a model that does not expect blocking operations. The authors of the study provided mutually recursive futures as an example---if two futures each attempted to read the value of the other, a deadlock would occur.

  \item A model might not guarantee safety or liveness by design, making composability impossible. Perhaps obviously, a model that does not guarantee safety or liveness would violate the safety guarantee of an outer model unless such a conflict was explicitly prevented.
\end{itemize}

The study noted that some types of bad interactions were already prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds. The authors noted that similar safeguards could prevent other negative interactions. However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models. Therefore, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.

\section{Unifying abstractions}

Other research into composable concurrency models has focused on the development of underlying abstractions capable of supporting a wide variety of models. Instead of programmers using composable ``building blocks'' to construct models, these abstractions would make use of those common elements at a lower level to improve performance and interoperability.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly. To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:

\begin{enumerate}
  \item Can the concept be reasonably implemented as a library? A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant virtual machine inclusion.

  \item Does the concept require runtime support to guarantee its semantics? Semantic guarantees may be enforced by a compiler but not the virtual machine. For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.

  \item Would runtime support benefit performance? Deeper knowledge of the concept may enable the virtual machine to better optimize. Some concepts, though, may require knowledge of the underlying hardware to realize any performance improvements; such concepts would not necessarily benefit from runtime support.

@@

\begin{itemize}
  \item Managed mutation: A concurrency model may impose rules on how some state can be modified; therefore, mutation must be able to be flexibly managed.

@@

  \item Ownership: An owning entity regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

  \item Leveled reflection: Reflection allows a program to examine its own structure and behavior at runtime. To guarantee safety, there needs to be a distinction between language-level reflection (reflection operations supported by the high-level language) and meta-level reflection (reflection operations on the low-level abstraction).

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/obmop.tikz}}
  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

@@

There are a rich variety of concurrency models that can be used to represent concurrent tasks, abstract over low-level details, and ensure that concurrent applications perform correctly. Recent research has attempted to identify ways in which these models can be safely used together, common features of these models, and underlying abstractions capable of representing many different models.

Future work will likely explore ways in which concurrency models can be decomposed and recomposed from common elements. This, in turn, would lead to the development of more generalized ways of describing concurrency models and perhaps additional unifying abstractions. As Marr and D'Hondt note, a ``silver bullet'' does not exist. However, there is certainly promise in searching for better ways to represent concurrent operations.

@@

Thanks to Elena Machkasova, K.~K. Lamberty, and Matthew Justin for their suggestions and feedback.
